---
title: "Destructive Power of Storms"
author: "Dave Lofte"
date: "Saturday, February 21, 2015"
output: html_document
---
### Synopsis
In this analysis we use data from the National Weather Service to try to determine which
weather events are the most destructive economically and harmful to human health.

##Processing the Data

###Reading in the data
We read in the data, and take a quick look using the str() function.  We notice that there are a few variables that are not necessary for our analysis.
```{r}
setwd("~/R/reproResearch/RepData_PeerAssessment2")

library(dplyr)
library(reshape2)
library(ggplot2)

url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if (!"storm.csv.bz2" %in%dir()){
    download.file(url, "storm.csv.bz2")}
stormData <- read.csv(bzfile("storm.csv.bz2"))
str(stormData)
```


Exploring over what time period most of the observations occur.
```{r}
stormData$year <- as.numeric(format(as.Date(stormData$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))

hist(stormData$year, breaks = 30)
```


The number of observations increases sharply after 1990ish, so we restrict our attention to dates after 1990
```{r}
trimStorm <- stormData[stormData$year >= 1990,]
```
We now separate the data into two different dataframes.
```{r}
health <- select(trimStorm, EVTYPE, FATALITIES, INJURIES)

economic <- select(trimStorm, EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) 
```

### Economic processing
Convert propdmg and cropdmg to same units using the propdmgexp and cropdmgexp variables.  I created two conversion vectors to facilitate this process. The conversion values were found investigating the storm data manual and investing some of the original data, the most intersting and useful fact discovered was that the numeric values in the 'exp' variables were overflows from the propdmg and cropdmg variables, thus they represent an extra decimal place for the original data which is why they are converted to "10".

```{r}
replaced <- c("H",  "K",  "M",     "B",        "-", "+", "0", "1", "2", '3', '4', '5', '6', '7', '8', "", "?" )
replacer <- c( 100, 1000, 1000000, 1000000000, 0,   1,    10,  10,  10,  10,  10,  10,  10,  10,  10,  0,  0 )

economic$PROPDMGEXP <- as.character(toupper(economic$PROPDMGEXP))
economic$CROPDMGEXP <- as.character(toupper(economic$CROPDMGEXP))
```
We use a loop subsetting our data using the replaced coversion vector and replacing with the associated value from the replacer vector.  Lastly we mutate PROPDMG and CROPDMG multiplying by the CROPDMGEXP and PROPDMGEXP.

All this work allows to have comparable values in the PROPDMG and CROPDMG variables.
```{r}

len <- length(replaced)
for (idx in 1:len){
    economic$PROPDMGEXP[economic$PROPDMGEXP == replaced[idx]] <-replacer[idx]
}

for (idx in 1:len){
    economic$CROPDMGEXP[economic$CROPDMGEXP == replaced[idx]] <- replacer[idx]
}
economic <- transform(economic, PROPDMGEXP =as.numeric(PROPDMGEXP), CROPDMGEXP = as.numeric(CROPDMGEXP))
economic <- mutate(economic, PROPDMG = PROPDMG*PROPDMGEXP, CROPDMGEXP = CROPDMG*CROPDMGEXP)

```

We then summarise the data averaging the cost of property damage and crop damage by  grouping the data by the variable EVTYPE whose values are different types of weather events.
```{r}
econo <- melt(economic,id.vars= "EVTYPE",measure.vars = c("PROPDMG","CROPDMG"),variable.name= "DamageType", value.name = "Cost")

#damages<- aggregate(econo$Cost, list(EVTYPE = econo$EVTYPE, DamageType = econo$DamageType), mean)

damage <- head(arrange(aggregate(econo$Cost, list(EVTYPE = econo$EVTYPE, DamageType = econo$DamageType), mean), desc(x)),10)

```

### Health Processing
We then summarise the data averaging the number of fatalities and injuries grouping the data by the variable EVTYPE whose values are different types of weather events.
```{r}

casual <- melt(health, id.vars= "EVTYPE" )

PopulationHealth <- arrange(aggregate(casual$value, list(EVTYPE = casual$EVTYPE, Casualty = casual$variable), mean),  desc(x))
    
pophealth <- head(PopulationHealth,20)
```

##Results
The chart below shows the average amount of economic dammage caused by the top 10 damaging weather events. It turns out that the top 10 damaging weather events all damaged property.  We can see that the most destructive weather events are Tornados, Tropical wind storm, and Hail events followed by Heavy Rain and sever weather events.  These two category of weather events caused the lions share of the damage.

```{r}
qplot(EVTYPE, x, data = damage, geom = "bar", stat = "identity", xlab = "", ylab = "Average Cost") +
   facet_grid(.~DamageType) +
theme(text = element_text(size= 15), axis.text.x = element_text(angle = 90, vjust = 0, hjust=1))
```


The graph below shows the average number of casualties(fatalities/injuries) for the top 10 most harmful weather events.  We can see that overall there are many more injuries form the result of weather events than deaths. The weather event most likely to cause non fatal injuries are surprisingly heatwaves.  The weather event that results in the most events are Tornadoes, Tropical wind storms, and Hail events.  
```{r}
qplot(EVTYPE, x, data = pophealth, geom = "bar", stat = "identity", xlab = "", ylab = "Average Casualties")+
    facet_grid(.~Casualty)+
    theme(text = element_text(size = 15), axis.text.x = element_text(angle = 90, vjust = 0, hjust = 1))
```

